---
title: "Statistical Inference Part #1 Simulation Exercise"
author: "Valentina Scipione"
date: "24 May 2015"
output: pdf_document
---

# Simulations

The exponential distribution is simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. You will investigate the distribution of averages of 40 exponentials with 1000 simulations.

We begin with 1000 simulated averages of 40 exponentials.

```{r}
# Exponential Distribution
set.seed(3)  # Set seed
n <- 40  # number of exponential random variables
lambda <- 0.2  # lambda for all simulations
nsim <- 1000  # number of simulated averages

mns = NULL
vars = NULL
for (i in 1 : nsim) {
  expd <- rexp(n, lambda)
  mns  <- c(mns, mean(expd))
  vars <- c(vars, var(expd))      
}

mean_sample <- mean(mns)  # mean of distribution of averages of 40 exponentials
mean_theoretical <- 1/lambda  # mean from analytical expression
```

```{r}
summary(mns)
summary(vars)

mean_sample
mean_theoretical
```

# Results

**1. Show where the distribution is centered at and compare it to the theoretical center of the distribution**

## Plot of the Distribution of 1000 collections of averages of 40 exponentials

```{r}
# Plot the distribution of 1000 collections of averages of 40 exponentials
library(ggplot2)
g1 <- qplot(mns,  fill=I("blue"), color=I("white"), geom="histogram", xlab="Mean", binwidth=0.2, xlim=c(1,9),
            main="Distribution of 1000 collections of averages of 40 exponentials") 
g1 <- g1 + geom_vline(xintercept = mean_sample, color="red") 
g1 <- g1 + geom_text(mapping=aes(x=mean_sample, y=110, label=paste("sample mean=",round(mean_sample,3))), size=4, vjust= 1, hjust=-0.1)
g1 
```

**2. Show how variable it is and compare it to the theoretical variance of the distribution**

```{r}
# Plot the distribution of variances
vars_sample <- mean(vars)
vars_theoretical <- (1/lambda)^2 #variance= std^2
vars_sample
vars_theoretical
g2 <- qplot(vars,  fill=I("blue"), color=I("white"), geom="histogram", binwidth=2, xlab="Variance of 40 Exponentials",
            main="Distribution of 1000 variances of collections of 40 exponentials") 
g2 <- g2 + geom_vline(xintercept = vars_sample, color="red") 
g2 <- g2 + geom_text(mapping=aes(x=vars_sample, y=130, label=paste("sample variance=",round(vars_sample,2))), size=4, hjust= -0.1)
g2 

# Compare with the distribution of 1000 exponentials
expd <- rexp(nsim, lambda)
expd_mean = mean(expd)
expd_var = var(expd)
g3 <- qplot(expd,  fill=I("blue"), color=I("white"), geom="histogram", xlab="Value", binwidth=1,
            main="Distribution of 1000 exponentials") 
g3 <- g3 + geom_vline(xintercept = expd_mean, color="red") 
g3 <- g3 + geom_text(mapping=aes(x=expd_mean, y=180, label=paste("sample mean=",round(expd_mean,3))), size=4, hjust= -0.1, vjust=1)
g3

library(grid)
library(gridExtra)
grid.arrange(g1, g3, ncol=2)
```
```

**3. Show that the distribution is approximately normal**

```{r}
# use qqplot and qqline to compare the distribution of averages of 40 exponentials to a normal distribution
qqnorm(mns, col="blue")
qqline(mns, col = 2)
```

